---
title: "Module One"
author: "Claire Fields"
date: "2026-05-20"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r workshop1}
source("workshop1.R")
```

## Workshop 2: Data Visualisations

# Install Packages
```{r}
library("tidyverse")
library("ggplot2")
```

```{r graphing template}
#ggplot(data=<DATA>) +
  #<GEOM_FUNCTION>(mapping=aes(<XYMAPPINGS))
# geom is how you represent your data
#point, line, smooth
#aes indicates the aesthetics, include x and y here
#add colour = z to map z to colour (changes colour based on class)
#alpha=transparency, colour, size, shape

#updated at the end of workshop 2:
#ggplot(data = <DATA>) + 
  #<GEOM_FUNCTION>(
     #mapping = aes(<MAPPINGS>),
     #stat = <STAT>, 
     #position = <POSITION>
 # ) +
#  <FACET_FUNCTION>
```

```{r practice plots}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ, y=hwy, shape=class))

#a question to try out
ggplot(data=mpg)+
  geom_point(mapping=aes(x=displ,y=hwy, colour=displ<5))
```

```{r facet and panel plots}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl)
#facet_grid   separate variables by ~ if you only want one use a .
#nrow is number of rows, ncol is number of columns
```

```{r lines}
ggplot(data = mpg) + 
  # geom_point(mapping=aes(x=displ,y=highway)) #points look bad
  geom_smooth(mapping = aes(x = displ, y = hwy))

 ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, colour = drv), show.legend = FALSE)
#can add group= to group the data but it doesnt add a legend,
 
 ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
 #not ideal because it increase the chance of a mistype if you go to change things
 
#instead:
 ggplot(data = mpg, mapping=aes(x=displ,y=hwy)) + #makes mappings global to the plot
  geom_point()+
   geom_smooth()
 # you can individually add aesthetics to each geom
 
 ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)

 #filter means you are only showing subcompact cars on the line
 
 #these should be the same
 ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
#and they are!
```

```{r transformations}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))


#override a default stat
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")

#override into proportions
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))

#improve transparency by using stat_summary
ggplot(data = diamonds) + 
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  )
```

```{r Aesthetic Adjustments}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) #colour just changes the outline
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) #fill actually fills in the shape
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity)) #breaks the cut into clarity

#position adjustments: identity = raw data, fill = changes height, dodge = forces ggplot2 to not put things on top of eachother

#change transparancy with alpha
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#color outlines with no fill
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")


#position fill means that you make all the stacks the same height, showing more of a percentage instead of actual count
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")

# dodge moves them side by side
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")

# jitter slightly move the points
ggplot(data = diamonds) + 
  geom_point(mapping = aes(x = cut, y = clarity), position = "jitter")
```



## Workshop 3: Building a tool box

```{r FOR LOOPS}
#for (value in that){
 # this
#}

#this is telling r to print "one run" for every value in c
for (value in c("My", "first", "for", "loop")){
  print("one run")
  }

#this is telling r to print the value of each entry in c
for (value in c("My", "second", "for", "loop")){
  print(value)  #can change print to anything
  }

my_list<- 1:50 
print(my_list)
str(my_list)

#make a for loop
for(i in my_list){   #heading of the for loop
  result <- i
  out <- paste0("I am printing loop number: ", i, ".")
  print(out)
  Sys.sleep(0.5) #this slows it down for us to see
}
print("For loop complete!")
print("Program complete")

#more complex
print("This for loop will add 500 to each iteration")
add_me<-500
for(i in my_list){
  result<- i + add_me
  out<- paste0("Result of iteration: ", i, "=", result, ".")
              print(out)
}
print("Complete!")

```

```{r NESTED FOR LOOPS}
for(i in c(1,2,3,4,5)){  #for every iteration in the list
  print(paste("I am base loop number:", i))   #print this
 # Sys.sleep(0.8)  #slows down

  for(j in c(1,2,3)){     #for every iteration in c
    print(paste("...I am inside loop number:",j))   #print this
    #Sys.sleep(0.8)
  }  # this closes the loop, if you move it up before the next loop then it wouldn't be nested and just one after another
}    
```

### Github Workflow

Edit → Save → Stage → Commit → Pull → Push

## Workshop 4: Developer's Toolbox and AI 

```{r}
library(ellmer)
library(chattr)
```

```{r store api key}
usethis::edit_r_environ()
usethis::edit_r_profile()
#have to save and restart r after each of these actions. stores them in the environment and profile instead of my actual code
```

```{r practicing chattr}
# ctrl + o opens chat
# prompt:show me the code to make a scatter plot showing the relationship between size and clarity of diamonds in the diamonds base data set 

# Code:
library(ggplot2)

ggplot(data = diamonds, aes(x = carat, y = clarity)) +
  geom_point()
```

```{r practicing inline copilot}
#Load tidyverse and palmerpenguins libraries
library(tidyverse)
library(palmerpenguins)


#Create a ggplot scatter plot using this penguins data
#plot bill_len on the x axis and bill_dep on the y axis
#color points by species and use geom_point with size 3
#add a minimal theme and clean labels
ggplot(data = penguins, aes(x = bill_len, y = bill_dep, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Bill Length vs Bill Depth of Penguins",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")


```

```{r}
#option A
# please create a boxplot that compares flipper length across different species, faceted by island, use custom color palettes
ggplot(data = penguins, aes(x = species, y = flipper_len, fill = island)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Biscoe" = "lightblue", "Dream" = "lightgreen", "Torgersen" = "lightpink")) +
  theme_minimal() +
  labs(title = "Flipper Length Comparison Across Penguin Species",
       x = "Species",
       y = "Flipper Length (mm)",
       fill = "Island")

# option b
#generate a boxplot comparing flipper_len across different species, faceted by island, using custom color palettes. i would like the island names to be along the x axis and no labels for the species other than the fill
ggplot(data = penguins, aes(x = island, y = flipper_len, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ species) +
  scale_fill_manual(values = c("Adelie" = "lightblue", "Chinstrap" = "lightgreen", "Gentoo" = "lightpink")) +
  theme_minimal() +
  labs(title = "Flipper Length Comparison Across Penguin Species",
       x = "Island",
       y = "Flipper Length (mm)",
       fill = "Species")
# option c
ggplot(data = penguins, aes(x = species, y = flipper_len, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ island) +
  scale_fill_manual(values = c("Adelie" = "lightblue", "Chinstrap" = "lightgreen", "Gentoo" = "lightpink")) +
  theme_minimal() +
  labs(title = "Flipper Length Comparison Across Penguin Species",
       x = "Species",
       y = "Flipper Length (mm)",
       fill = "Species")
# option d
ggplot(data = penguins, aes(x = island, y = flipper_len, fill = species)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Adelie" = "lightblue", "Chinstrap" = "lightgreen", "Gentoo" = "lightpink")) +
  theme_minimal() +
  labs(title = "Flipper Length Comparison Across Penguin Species",
       x = "Island",
       y = "Flipper Length (mm)",
       fill = "Species")
```

I tried a couple different plots using different levels of copilot to help me out. I think option D is the best, even though it is not technically faceted. I think it is more visually appealing and easier to read than the other options. 
# 4.3 Conditional Logic
The classic base R conditional structure screens a single value at a time. Its basic anatomy uses round brackets () for the logical condition and curly braces {} for the action if that condition is true. Base r basic conditional logic only works for one value, can't use vectors.
```{r practice}
# Simple Example: Checking Sea Surface Temperature (SST)
sst <- 30.2

if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
} else {
  print("SST remains within baseline parameters.")
}
```

Use tidyverse to evaluate columns
```{r}
# Sample coral data
coral_monitoring <- tibble(
  site= c("A", "B", "c"),
  depth_m = c(12, 35, 8)
  )

#classify depth using if_else
#coral_monitoring <- 
  coral_monitoring %>%
  mutate(zone=if_else(depth_m>30,"Deep Reef", "Shallow Reef"))

# mutate: add a new column
# %>%: pipe one variable/object to a function




```



